### 2021-06-03 JSS/using_drtmle_jss.Rmd
update manuscript

In many fields researchers are interested in assessing the population-level
effect of a particular treatment on an outcome of interest. The treatment might
correspond to a drug, a potentially harmful exposure exposure, or a policy
intervention. Often, the treatment may not be randomized due to ethical or
logistical reasons. This has sparked great interest in developing statistical
methodology to estimate population-level effects from observational data.
Estimation of these effects often requires accounting for putative confounding
factors, that is, factors related to whether a participant receives treatment
and to the participant's outcome. In many settings, researchers measure many
potential confounders, for example using administrative databases or genetic
sequencing technology. Due to the curse-of-dimensionality, common statistical
estimation techniques are not feasible for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated,
estimates of the population-level effect of a treatment may have large bias.

This has sparked an interest in using adaptive estimation techniques from the
machine learning literature to control for high-dimensional confounders when
estimating treatment effects. However, facilitating proper inference (e.g.,
confidence intervals and hypothesis tests) about estimates of effects based on
adaptive estimators is challenging. In particular, standard causal inference
techniques, including both G-computation (GCOMP) estimators
[@robins:1986:mathmod] and inverse probability of treatment weighted (IPTW)
estimators [@horvitzthompson:1952:jasa, @robins:2000:epi], fail to yield valid
inference for common estimators of effects. More complex proposals have been
developed that *are capable* of utilizing adaptive methods to control for
confounding while still yielding valid inference. These techniques include
augmented inverse probability of treatment estimation (AIPTW)
[@robins:1994:jasa] and targeted minimum loss-based estimation (TMLE)
[@vdlrubin:2006:ijb]. Under assumptions, these estimators have limiting normal
distributions with estimable variance. The asymptotic variance estimates then
serve as the basis for constructing confidence intervals and hypothesis tests.

@@

Suppose the observed data consist of $n$ independent and identically
distributed copies of the random variable $O = (W,A,Y) \sim P_0$. Here, $W$ is
a vector of baseline covariates, $A\in\{0,1\}$ is a binary (for the sake of
simplicity) treatment, $Y$ is an outcome, and $P_0$ is the true data-generating
distribution. The parameters of interest are \begin{equation}
\psi_0(a) = E_0\{\bar{Q}(a,W)\} = \int \bar{Q}_0(a,w) dQ_{W,0}(w) \ , \
\mbox{for} \ a = 0,1 \ ,
\label{parameterDef}
\end{equation}
where $\bar{Q}_0(a,w)=E_0 \left(Y \mid A=a, W=w\right)$ is the so-called
outcome regression and $Q_{W,0}(w)=Pr_0\left(W\leq w\right)$ is the distribution
function of the covariates. The value $\psi_0(a)$ represents the marginal
(i.e., population-level)treatment-specific average outcome, hereafter referred
to as the *marginal mean* under treatment $A=a$. The marginal effect of the
treatment on the outcome can be assessed by comparing marginal means under
different treatment levels. For example, the average treatment effect is often
defined as $\psi_0 := \psi_0(1) - \psi_0(0)$. Under additional assumptions,
these parameters have richer interpretations as mean counterfactual outcomes
under the different treatments and contrasts between these means define causal
effects.

@@

An estimator of $\psi_0(a)$ may be motivated directly by (1). Suppose we have
available an estimate of the OR, $\bar{Q}_n$, and an estimate of the
distribution function of baseline covariates, $Q_{W,n}$. An estimator of
$\psi_0(a)$ may be obtained by plugging these into (1), $\psi_n(a) = \int
\bar{Q}_0(a,w) dQ_{W,n}(w).$ Often the empirical distribution function of
covariates is used so that this estimator can be equivalently written as
\begin{equation*}
\psi_{n,GCOMP}(a) = \frac{1}{n}\sum\limits_{i=1}^n \bar{Q}_n(a, W_i) \ .
\end{equation*}
This estimator is commonly referred to as the G-computation (GCOMP) estimator.
Inference for GCOMP estimators based on parametric OR estimators may be
facilitated through the nonparametric bootstrap. However, if adaptive
approaches are used to estimate the OR, inference may not be available without
further modification to the estimator.

@@

The main function of the package is the eponymous `drtmle` function. This
function estimates the treatment-specific marginal mean for user-specified
levels of a discrete-valued treatment and computes a doubly-robust covariance
matrix for these estimates. The `drtmle` function implements either the
estimators of \citet{vdl:2014:ijb} (if `reduction = "bivariate"`) or the
improved estimators of \citet{benkeser:2017:biometrika} (`reduction = "univariate"`). The
package includes utility functions for combining these estimates into estimates
of average treatment effects, as well as other contrasts, as discussed below.

The main data arguments for `drtmle` are `W, A, Y`, which, as above, represent
the covariates, a discrete-valued treatment, and an outcome, respectively.
Missing data are allowed in `A` and `Y`; this is discussed further in a section
below. Beyond the data arguments, it is necessary to specify how to estimate
each of the OR, PS, R-OR, and R-PS (or R-PS1, R-PS2). The package provides
flexibility in how these regression parameters may be estimated. Once the user
has specified the data arguments and how to estimate each regression parameter,
the function proceeds as follows:
  1. estimate the OR via the internal `estimateQ` function;
  2. estimate the PS via the internal `estimateG` function;
  3. estimate reduced-dimension regression via `estimateQrn` and
     `estimategrn` functions;
  4. iteratively modify initial estimates of the OR and PS via the `fluctuateQ`
      and `fluctuateG` functions until equations (5)-(7) are approximately
      solved;
  5. compute the TMLE estimates and covariance estimates based on the final OR.
We refer interested readers to \citet{benkeser:2017:biometrika} for a theoretical
derivation of how the iterative modification of initial OR and PS is performed.

@@

Generalized linear models are perhaps the most popular means of estimating
regression functions. Because of this legacy, these estimators have been
included as an option for estimating the regression parameters. The user
specifies the right-hand-side of a regression formula as a character in the
`glm_Q`, `glm_g`, `glm_Qr`, and `glm_gr` options to estimate the OR, PS, R-OR,
and R-PS. These regressions are fit by calling the `glm` function in base R.
Thus, the model specification should be able to be parsed as a valid formula,
as required by `glm`. Recall that the R-OR parameter involves a regression onto
the estimated propensity. Thus, the formula that is input via `glm_Qr` should
assume data with a single column named `"gn"`. Similarly, the R-PS1 and R-PS2
that are computed for the estimators of \citet{benkeser:2017:biometrika} (the default
estimators, computed whenever `reduction = "univariate"`) involve regressions
onto the estimated outcome regression. Thus, the formula `glm_gr` should assume
data with a single column named `"Qn"` if `reduction = "univariate"`. If
instead, the estimators of \citet{vdl:2014:ijb} are preferred, we can set the
option `reduction = "bivariate"` and input a `glm_gr` formula that assumes data
with two columns named `"Qn"` and `"gn"`.

The code below illustrates how generalized linear models can be used to
estimate the regression parameters. In this call to `drtmle`, we specify
`family = binomial()` so that logistic regression is used to estimate the OR.
We specify `stratify = FALSE` to indicate that we wish to fit a single OR to
estimate $\bar{Q}_0(A,W)$, as opposed to fitting two separate regressions to
estimate $\bar{Q}_0(1,W)$ and $\bar{Q}_0(0,W)$. If the former fitting is
specified, then the OR regression is fit using a data frame that contains `W`
and `A`. Thus, both `colnames(W)` and `"A"` may appear in `glm_Q`. However, if
`stratify = TRUE` then the OR model is fit separately in observations with $A =
1$ and $A = 0$. In this case, the `glm_Q` option may not include `"A"`. We
illustrate non-stratified regression for both the univariate and bivariate
reductions.

@@

glm_fit_biv <- drtmle(W = W, A = A, Y = Y, family = binomial(),
                      glm_g = "W1 + W2", glm_Q = "W1 + W2*A",
                      glm_gr = "Qn", glm_Qr = "gn", stratify = FALSE,
                      reduction = "bivariate")
glm_fit_biv
```

@@

```{r, echo = FALSE, message = FALSE}
library(SuperLearner)
```
Super learning is a loss-based ensemble machine learning technique
[@vdl:2007:statapp] that is a generalization of stacking and stacked regression
[@wolpert:1992:nnet]breiman:1996:ml}. The user specifies many potential
candidate regression estimators, referred to as a *library* of candidate
estimators, and the super learner algorithm estimates the convex combination of
regression estimators that minimizes $V$-fold cross-validated risk based on a
user-selected loss function. Oracle inequalities show that the super learner
estimate performs essentially as well as the unknown best convex combination of
the candidate estimators [@vdvaart:2006:statdec]. Thus, the methodology may be
seen as an asymptotically optimal way of performing estimator selection.
Practically, the super learner provides the best chance for obtaining a
consistent estimator for both the OR and PS, and thereby an efficient estimator
of the marginal means of interest.

Super learning is implemented in the R package `SuperLearner`
[@SuperLearnerPackage] and `drtmle` provides the option to utilize super
learning for estimation of the OR, PS, and reduced-dimension regressions via
the `SL_Q`, `SL_g`, `SL_Qr`, and `SL_gr` options. When calling `drtmle` either
the generalized linear model options for regression modeling or the super
learner options should be invoked. If both are called, the function will
default to the super learner option. The inputs to the super learner options
are passed to the `SL.library` argument of the `SuperLearner` function of the
`SuperLearner` package. We refer readers to the `SuperLearner` help file for
information on how to properly specify a super learner library. In the code
below, we illustrate a call to `drtmle` with a relatively simple library. For
the OR and PS, the library includes a main-terms generalized linear model (via
the function `SL.glm` from `SuperLearner`) and an unadjusted
generalized linear model (`SL.mean`). For the reduced dimension regressions we
use a super learner library with a main terms generalized linear model
(`SL.glm`) and adaptive regression splines (`SL.earth`, which utilizes the `earth`
package [@earthPackage]).

@@

Some users may find it unsettling that when applying the super learner in real data 
analysis is that the results are dependent on the random seed that is set. To remove 
this dependence, one can instead average results over repeated super learner runs; more 
repeats will lead to less dependence on the random seed. There are generally two approaches
that can be used: (i) averaging on the scale of the nuisance parameters and (ii) averaging
on the scale of the point estimates.

@@

For(ii), suppose for example that we have $n_{SL}$ TMLE (similarly, AIPTW) estimates 
$\psi_{n,j}(1)$ of $\psi_0(1)$, each obtained based on one (or more) super learner of the OR and/or PS.
We can take the average over these point estimates, \[
\psi_n(1) = \frac{1}{n_{SL}} \sum_{j=1}^{n_{SL}} \psi_{n,j}(1)
\]
as our estimate of $\psi_{0}(1)$.

@@

### Outcome-adaptive propensity score

If the outcome-adaptive PS is desired, the user may specify this via 
the `adapt_g` option. In this case, the additional targeting steps needed for doubly
robust inference are "turned off" (i.e., `guard = NULL`), since, as discussed above,
this estimator is *not* doubly robust. Note that when the `adapt-g = TRUE` the data
used for the PS estimation will be a `data.frame` with column names
`QaW` where `a` takes all the values specified in `a_0`. For example if `a_0 = c(0,1)`,
then the PS estimation data frame will have two columns called `Q0W`
and `Q1W`. This may be relevant for users writing their own `SuperLearner` wrappers
or those who make use of the `glm_g` setting, as illustrated below.

@@

Obtaining fully cross-validated standard errors can be fairly time intensive when super learner is
used to estimate the OR and PS. In `drtmle` we give an option to compute standard error estimates that
are *partially cross-validated* and do not require any additional fitting. Here we describe briefly 
how these estimates are obtained and show calls to `drtmle` to obtain these estimates.

@@

>>>>

In many fields researchers are interested in assessing the population-level
effect of a particular treatment on an outcome of interest. The treatment might
correspond to a drug, a potentially harmful exposure, or a policy
intervention. Often, the treatment cannot be randomized due to ethical or
logistical reasons. This has sparked great interest in developing statistical
methodology to estimate population-level effects from observational data.
Estimation of these effects often requires accounting for putative confounding
factors, that is, factors related to whether a participant receives treatment
and to the participant's outcome. In many settings, researchers measure many
potential confounders, for example using administrative databases or genetic
sequencing technology. Due to the curse-of-dimensionality, common statistical
estimation techniques are not feasible for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated,
estimates of the population-level effect of a treatment may have large bias.

This has sparked an interest in using adaptive estimation techniques from the
machine learning literature to control for high-dimensional confounders when
estimating treatment effects. However, facilitating proper inference (e.g.,
confidence intervals and hypothesis tests) about estimates of effects based on
adaptive estimators is challenging. In particular, standard causal inference
techniques, including both G-computation (GCOMP) estimators
[@robins:1986:mathmod] and inverse probability of treatment weighted (IPTW)
estimators [@horvitzthompson:1952:jasa, @robins:2000:epi], fail to yield valid
inference for common estimators of effects. More complex proposals have been
developed that *are capable* of utilizing adaptive methods to control for
confounding while still yielding valid inference. These techniques include
augmented inverse probability of treatment estimation (AIPTW)
[@robins:1994:jasa] and targeted minimum loss-based estimation (TMLE)
[@vdlrubin:2006:ijb]. Under assumptions, these estimators have limiting normal
distributions with estimable variance. Estimates of the asymptotic variance then
serve as the basis for constructing confidence intervals and hypothesis tests.

@@

Suppose the observed data consist of $n$ independent and identically
distributed copies of the random variable $O = (W,A,Y) \sim P_0$. Here, $W$ is
a vector of baseline covariates, $A\in\{0,1\}$ is a binary (for the sake of
simplicity) treatment, $Y$ is an outcome, and $P_0$ is the true data-generating
distribution. The parameters of interest are \begin{equation}
\psi_0(a) = E_0\{\bar{Q}(a,W)\} = \int \bar{Q}_0(a,w) dQ_{W,0}(w) \ , \
\mbox{for} \ a = 0,1 \ ,
\label{parameterDef}
\end{equation}
where $\bar{Q}_0(a,w)=E_0 \left(Y \mid A=a, W=w\right)$ is the so-called
outcome regression and $Q_{W,0}(w)=Pr_0\left(W\leq w\right)$ is the distribution
function of the covariates. The value $\psi_0(a)$ represents the marginal
(i.e., population-level)treatment-specific average outcome, hereafter referred
to as the *marginal mean* under treatment $A=a$. The marginal effect of the
treatment on the outcome can be assessed by comparing marginal means under
different treatment levels. For example, the average treatment effect is often
defined as $\psi_0 := \psi_0(1) - \psi_0(0)$. Under additional assumptions,
these parameters have interpretations as mean counterfactual outcomes
under the different treatments and contrasts between these means define causal
effects.

@@

An estimator of $\psi_0(a)$ may be motivated directly by (1). Suppose we have
available an estimate of the OR, $\bar{Q}_n$, and an estimate of the
distribution function of baseline covariates, $Q_{W,n}$. An estimator of
$\psi_0(a)$ may be obtained by plugging these into (1), $\psi_n(a) = \int
\bar{Q}_0(a,w) dQ_{W,n}(w).$ Typically, the empirical distribution function of
covariates is used so that this estimator can be equivalently written as
\begin{equation*}
\psi_{n,GCOMP}(a) = \frac{1}{n}\sum\limits_{i=1}^n \bar{Q}_n(a, W_i) \ .
\end{equation*}
This estimator is commonly referred to as the G-computation (GCOMP) estimator.
Inference for GCOMP estimators based on parametric OR estimators may be
facilitated through the nonparametric bootstrap. However, if adaptive
approaches are used to estimate the OR, inference may not be available without
further modification to the estimator.

@@

The main function of the package is the eponymous `drtmle` function. This
function estimates the treatment-specific marginal mean for user-specified
levels of a discrete-valued treatment and computes a doubly-robust covariance
matrix for these estimates. The function implements either the
estimators of \citet{vdl:2014:ijb} (if `reduction = "bivariate"`) or the
improved estimators of \citet{benkeser:2017:biometrika} (`reduction = "univariate"`). The
package includes utility functions for combining these estimates into estimates
of average treatment effects, as well as other contrasts, as discussed below.

The main data arguments for `drtmle` are `W, A, Y`, which, as above, represent
the covariates, a discrete-valued treatment, and an outcome, respectively.
Missing data are allowed in `A` and `Y`, as discussed below.
Beyond the data arguments, it is necessary to specify how to estimate
each of the OR, PS, R-OR, and R-PS (or R-PS1, R-PS2). The package provides
flexibility in how these regression parameters may be estimated. Once the user
has specified the data arguments and how to estimate each regression parameter,
the function proceeds as follows:
  1. estimate the OR via the internal `estimateQ` function;
  2. estimate the PS via the internal `estimateG` function;
  3. estimate reduced-dimension regression via `estimateQrn` and
     `estimategrn` functions;
  4. iteratively modify initial estimates of the OR and PS via the `fluctuateQ`
      and `fluctuateG` functions until equations (5)-(7) are approximately
      solved;
  5. compute the TMLE estimates and covariance estimates based on the final OR.
We refer interested readers to \citet{benkeser:2017:biometrika} for a theoretical
derivation of how the iterative modification of initial OR and PS is performed.

@@

Generalized linear models are perhaps the most popular means of estimating
regression functions. Because of this legacy, these estimators have been
included as an option for estimating the regression parameters. The user
specifies the right-hand-side of a regression formula as a character in the
`glm_g`, `glm_Q`, `glm_Qr`, and `glm_gr` options to estimate the OR, PS, R-OR,
and R-PS. These regressions are fit by calling the `glm` function in base R.
Thus, the model specification should be able to be parsed as a valid formula,
as required by `glm`. For the PS formula specified by `glm_g`, `colnames(W)` are 
available to use in the formula. For the OR (`glm_Q`) available names
of variables to be used in the formula depends on the `stratify` option.
If `stratify = FALSE`, a single OR is fit to
estimate $\bar{Q}_0(A,W)$. In this case, the OR regression is fit passing a `data.frame` into the `data` argument of `glm` that contains `W` and `A`. Thus, both `colnames(W)` and `"A"` may appear in the formula `glm_Q`. If `stratify = TRUE`, regressions are fit stratified by `A` to separately 
estimate $\bar{Q}_0(1,W)$ and $\bar{Q}_0(0,W)$. In this case, the `glm_Q` option 
may not include `"A"`.

To understand the variable names available for R-OR, recall that the parameter 
involves a regression onto the estimated propensity. Thus, the formula that is 
input via `glm_Qr` should
assume a data with a single column, `"gn"`. 
Similarly, the R-PS1 and R-PS2
that are computed for the estimators of \citet{benkeser:2017:biometrika} (the default
estimators, computed whenever `reduction = "univariate"`) involve regressions
onto the estimated outcome regression. Thus, the formula `glm_gr` should assume
data with a single column named `"Qn"` if `reduction = "univariate"`. If
instead, the estimators of \citet{vdl:2014:ijb} are preferred (`reduction = "bivariate"`) the 
`glm_gr` formula assumes data with two columns, `"Qn"` and `"gn"`.

The code below illustrates how generalized linear models can be used to
estimate the regression parameters. In this call to `drtmle`, we specify
`family = binomial()` so that logistic regression is used to estimate the OR.
We illustrate non-stratified regression for both the univariate (default) and bivariate
reductions.

@@

glm_fit_biv <- drtmle(W = W, A = A, Y = Y, family = binomial(),
                      glm_g = "W1 + W2", glm_Q = "W1 + W2*A",
                      glm_gr = "Qn", glm_Qr = "gn + Qn", stratify = FALSE,
                      reduction = "bivariate")
glm_fit_biv
```

@@

```{r, echo = FALSE, message = FALSE}
library(SuperLearner)
```
Super learning is a loss-based ensemble machine learning technique
[@vdl:2007:statapp] that generalizes stacking and stacked regression
\citep{wolpert:1992:nnet,breiman:1996:ml}. The user specifies many potential
candidate regression estimators, referred to as a *library* of candidate
estimators, and the super learner algorithm estimates the convex combination of
regression estimators that minimizes $V$-fold cross-validated risk based on a
user-selected loss function. Oracle inequalities show that the super learner
estimate performs essentially as well as the unknown best convex combination of
the candidate estimators [@vdvaart:2006:statdec]. Thus, the methodology may be
seen as an asymptotically optimal way of performing estimator selection.
Practically, the super learner provides the best chance for obtaining a
consistent estimator for both the OR and PS, and thereby an efficient estimator
of the marginal means of interest.

Super learning is implemented in the R package \pkg{SuperLearner}
[@SuperLearnerPackage] and `drtmle` provides the option to utilize super
learning for estimation of the OR, PS, and reduced-dimension regressions via
the `SL_Q`, `SL_g`, `SL_Qr`, and `SL_gr` options. When calling `drtmle` either
the generalized linear model options for regression modeling or the super
learner options should be invoked. If both are called, the function will
default to the super learner option. The inputs to the super learner options
are passed to the `SL.library` argument of the `SuperLearner` function of the
`SuperLearner` package. We refer readers to the `SuperLearner` help file for
information on how to properly specify a super learner library. In the code
below, we illustrate a call to `drtmle` with a relatively simple library. For
the OR and PS, the library includes a main-terms generalized linear model (via
the function `SL.glm` from `SuperLearner`) and an unadjusted
generalized linear model (`SL.mean`). For the reduced dimension regressions we
use a super learner library with a main terms generalized linear model
(`SL.glm`) and adaptive regression splines (`SL.earth`, which utilizes the `earth`
package [@earthPackage]).

@@

Some users may find it unsettling that when applying the super learner in real data 
analysis the results are dependent on the random seed that is set. To remove 
this dependence, one can average results over repeated super learner runs; more 
repeats will lead to less dependence on the random seed. There are generally two approaches
that can be used: (i) averaging on the scale of the nuisance parameters and (ii) averaging
on the scale of the point estimates.

@@

For (ii), suppose for example that we have $n_{SL}$ TMLE (similarly, AIPTW) estimates 
$\psi_{n,j}(1)$ of $\psi_0(1)$, each obtained based on one (or more) super learner of the OR and/or PS.
We can take the average over these point estimates, \[
\psi_n(1) = \frac{1}{n_{SL}} \sum_{j=1}^{n_{SL}} \psi_{n,j}(1)
\]
as our estimate of $\psi_{0}(1)$.

@@

@@

Obtaining fully cross-validated standard errors can be fairly time intensive when super learner is
used to estimate the OR and PS. In `drtmle` we give an option to compute standard error estimates that
are *partially cross-validated* and do not require any additional fitting beyond a standard super learner. 
Here we describe briefly how these estimates are obtained and show calls to `drtmle` to obtain these estimates.

@@

### Implementation of outcome-adaptive propensity score

If the outcome-adaptive PS is desired, the user may specify this via 
the `adapt_g` option. In this case, the additional targeting steps needed for doubly
robust inference are "turned off" (i.e., `guard = NULL`), since, as discussed above,
this estimator is *not* doubly robust. Note that when the `adapt_g = TRUE` the data
used for the PS estimation will be a `data.frame` with column names
`QaW` where `a` takes all the values specified in `a_0`. For example if `a_0 = c(0,1)`,
then the PS estimation data frame will have two columns called `Q0W`
and `Q1W`. This may be relevant for users writing their own `SuperLearner` wrappers
or those who make use of the `glm_g` setting.
